preciseTAD: A machine learning framework for precise 3D domain boundary prediction at base-level resolution

### 2021-01-21 BMC_Genome_Biology_submission/grubert_manuscript_figures_tables/manuscript/manuscript.Rmd
made edits to grubert version of manuscript

---
title: "preciseTAD: A machine learning framework for precise 3D domain boundary prediction at base-level resolution"
csl: styles.ref/genomebiology.csl
output:
  word_document:
    reference_docx: styles.doc/Arial_11_double_space_normal_margins.docx
  pdf_document: 
    extra_dependencies: ["float"]
bibliography: References/references.bib
---

@@

@@

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of the 3-dimensional (3D) structure of the human genome such as chromatin loops [@rao20143d], Topologically Associating Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At the kilobase scale, chromatin loops connect gene promoters with distal enhancers, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. At the megabase scale, TADs represent regions on the linear genome that are highly self-interacting. <!--Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa].--> Disruption of boundaries demarcating TADs and loops promotes cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Identifying the precise location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

@@

Another limitation among domain callers is that they disregard prior knowledge about functional genomic annotations associated with TAD and loop boundaries, referred hereafter as domain boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been identified as components of the loop extrusion model<!--, whereby DNA is extruded through cohesin rings forming chromatin loops--> [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Furthermore, distinct patterns of histone modifications have also been shown to be present at boundaries<!--including H3K36me3--> [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These genomic annotations are obtained using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). The resolution of ChIP-seq experiments is typically on the order of tens to hundreds of bases [@mendoza2009genome], well below the resolution of Hi-C data (tens of kilobases; 750 bp is the highest resolution of Hi-C data to date [@bonev2017multiscale]). Therefore, leveraging precisely mapped genomic annotations in a supervised machine learning framework enables more precise prediction of the locations of domain boundaries.

We present *preciseTAD*, an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on high-resolution chromatin state (BroadHMM), histone modification (HM), and transcription factor binding site (TFBS) data to predict low-resolution boundaries. We introduce a systematic pipeline for building the optimal boundary region prediction classifier. We found that spatial associations (linear distance) between boundaries and annotations perform best, transcription factor binding sites improve prediction performance, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and partitioning around medoids (PAM) to detect genome annotation-guided boundary regions and points at a base-level resolution. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that *preciseTAD* predictions are more enriched for known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD*-predicted boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in the domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on https://bioconductor.org/packages/preciseTAD.

@@

We developed a machine learning (ML) framework for determining the optimal set of data level characteristics to predict boundary regions of Topologically Associating Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We used Arrowhead-called TAD boundaries [@durand2016juicer] and published Peakachu-predicted loop boundaries [@salameh2019supervised] as ground truth. Data from GM12878 and K562 cell lines at 5-100 kb resolution (Arrowhead) and 10 kb resolution (Peakachu) were used (Additional file 1: Arrowhead Script, Additional file 2: Table S1<!--Supplementary Table 1-->). We next developed *preciseTAD*, an R package that leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base-level resolution from cell line-specific genome annotation data.

<!-- ![**Resolution-specific data construction and feature engineering for random forest modeling.** (A) The linear genome was binned into non-overlapping resolution-specific intervals using *shifted binning* (see Methods). The response vector **Y** was defined as 1/0 if a genomic bin overlapped/did not overlap with a TAD (or loop) boundary. (B) Four types of associations between bins (blue dashed lines) and genomic annotations (green shapes) were considered to build the predictor space, including Average Peak Signal (Signal), Overlap Counts (OC), Overlap Percent (OP), and $log_{2}$ distance (Distance).](Figures/figure1.png) -->

Boundary regions were defined as genomic bins containing a called boundary ($Y = 1$), while non-boundary regions were defined as bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Additional file 3: Table S2<!--Supplementary Table 3-->). The number of non-boundary regions highly outnumbered boundary regions. Such a disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address the class imbalance, we evaluated the effect of three resampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority class (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

|          Tool         |     Resolution    |     Total number of called TADs/chromatin loops    |     Total number of unique TAD/chromatin loop boundaries    |     Total number of genomic bins    |     Class imbalance    |
|:---------------------:|:-----------------:|:--------------------------------------------------:|:-----------------------------------------------------------:|:-----------------------------------:|:----------------------:|
|        Arrowhead      |       5   kb      |                         8052                       |                             15468                           |                535363               |           0.03         |
|     Arrowhead         |       10   kb     |                         7676                       |                             14253                           |                267682               |           0.05         |
|     Arrowhead         |       25   kb     |                         4670                       |                             8363                            |                107073               |           0.08         |
|     Arrowhead         |       50   kb     |                         2349                       |                             4224                            |                 53537               |           0.08         |
|     Arrowhead         |      100   kb     |                         1031                       |                             1883                            |                 26768               |           0.07         |
|        Peakachu       |       10   kb     |                        16185                       |                          21421                              |                267682               |        0.14            |
Table: Domain boundary data and class imbalance summaries across resolutions for Arrowhead and Peakachu on GM12878.

A total of 77 cell line-specific genomic annotations were used to build the predictor space. These included BroadHMM chromatin state data, histone modifications (HM), and transcription factor binding sites (TFBS) (Additional file 4: Table S3<!--Supplementary Table 2-->). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These included signal strength association (Signal), direct (OC), proportional (OP), and spatial ($log_2 + 1$ Distance) relationships (see Methods, Additional file 5: Figure S1<!--Supplementary Figure 1-->).

@@

<!-- ![**A machine learning framework for building domain boundary region prediction models.** Step 1 employs a range of feature engineering techniques to define the predictor matrix $A_{N \times (p)}$, where $N$ is the number of genomic bins, $p$ is the number of genomic annotations, $i$ is a holdout chromosome. The response vector $Y_N$ is defined as a boundary region ($Y=1$) if it overlaps with a genomic bin (else $Y=0$). Step 2 reserves the predictor-response matrix for the holdout chromosome $i$ as the test data. Step 3 applies a resampling technique to the training data to address the class imbalance. Step 4 trains the random forest model and performs 3-fold cross-validation to tune the mtry parameter. Finally, step 5 validates the model on the separate test data composed of the binned data from the holdout chromosome $i$ and evaluates model performance using balanced accuracy (BA).](Figures/figure2.png) -->

@@

Additionally, we found that using a distance-type predictor space yielded substantially higher balanced accuracy than the peak signal strength, overlap count, and overlap percent predictor types. As with class balancing techniques, this improvement was less evident at lower resolutions, with results consistent for K562 (Additional file 6: Figure S2A<!--Supplementary Figure 3A-->). Furthermore, 5 kb resolution genomic bins led to the optimal prediction for TAD boundary regions on both cell lines. Random forest models built on Peakachu-defined loop boundary regions yielded optimal prediction performance when using a distance-type predictor space, with SMOTE resampling only marginally outperforming RUS, for both cell lines (Additional file 6: Figure S2B, S2C<!--Supplementary Figure 3B,3C-->). Our results indicate that random under-sampling, distance-type predictors, and high-resolution Hi-C data provide the optimal set of data level characteristics for both TAD and chromatin loop boundary prediction.

<!-- ![**Determining optimal data level characteristics for building TAD boundary region prediction models on GM12878.** Averaged balanced accuracies are compared across resolution, within each predictor-type: Signal, OC, OP, and Distance, and across resampling techniques: no resampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Error bars indicate standard deviation from the mean performance across each holdout chromosome used for testing.](Figures/figure3.png) -->

@@

We hypothesized that the type of genomic annotations may also affect predictive performance. Using the established optimal settings (RUS, Distance, 5 kb/10 kb (Arrowhead/Peakachu ground truth) genomic bins), we built separate random forest models using histone modifications (HM), BroadHMM chromatin states (BroadHMM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, with results consistent for loop boundaries, on both cell lines (Figure 4A; Additional file 7: Figure S3A<!--Supplementary Figures 4A-->). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both GM12878 and K562.

<!-- ![**SMC3, RAD21, CTCF, and ZNF143 transcription factors accurately predict TAD and loop boundaries in GM12878.** (A) Barplots comparing performances of TAD (Arrowhead) and loop (Peakachu) boundary prediction models using histone modifications (HM), chromatin states (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent 1 standard deviation from the mean cross-validated accuracy across each holdout chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 most predictive chromosome-specific TFBSs. The columns represent the holdout chromosome excluded from the training data. Rows are sorted in decreasing order according to the columnwise average importance. (D) Receiver operating characteristic (ROC) curves and the corresponding average area under the curves (AUCs) when training and testing on GM12878 data (blue, Arrowhead ground truth; red, Peakachu ground truth) versus training on K562 and testing on GM12878 data (black, dashed). The curves represent the average sensitivities and specificities across each holdout chromosome. The shaded areas around each curve represent 1 standard deviation from the average.](Figures/figure4.png) -->

@@

We sought to further optimize our boundary region prediction models by implementing recursive feature elimination to avoid overfitting and selecting only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Additional file 7: Figure S3B<!--Supplementary Figures 4B-->). However, given that we trained our models on chromosome-specific data, the most significant annotations varied for each holdout chromosome. To determine transcription factors most important for boundary prediction across all chromosomes, we clustered the predictive importance (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found four transcription factors, CTCF, RAD21, SMC3, and ZNF143, being consistently predictive of TAD and loop boundaries (Figure 4C; Additional file 7: Figure S3C<!--Supplementary Figures 4C-->). We optimized our model by only considering these top four TFBS when building the random forest model, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the known molecular drivers of the loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line using the model pre-trained on data from another cell line (See Methods). We found that training and testing using Arrowhead ground truth TAD boundaries and genomic annotation data from the GM12878 cell line resulted in an average AUC=0.792 (Figure 4D). Interestingly, when training on the K562 cell line and testing on GM12878, the average AUC increased slightly to 0.795. Likewise, the average performance of models trained using Peakachu boundaries and genomic annotation data from the GM12878 cell line was comparable to models trained on K562-specific Peakachu boundaries and genomic annotations (Avg. AUC=0.881 and 0.874, respectively). These results were consistent when comparing training/testing strategies on K562 with training on GM12878 and testing on K562 data (Additional file 7: Figure S3D<!--Supplementary Figure 4D-->). The average ROC curves were found to be within 1 standard deviation of each other, suggesting that a model trained on data from one cell line performs well when using the data from another cell line, allowing for the opportunity to predict boundaries for cell lines that do not currently have Hi-C data available.

@@

Using our optimally built random forest model trained on Arrowhead/Peakachu boundaries, we attempted to predict boundaries at base-level resolution. Intuitively, instead of bin-level annotations, the predictor-response space was built on a base-level, that is, each base was associated with a genomic annotation. The model trained on a bin-level space was then applied on a base-level space to predict each base's probability of being a boundary. Our method, referred to as *preciseTAD*, uses density-based spatial clustering (DBSCAN) and partitioning around medoids (PAM) to predict the location of boundary regions and summit points (highest boundary probability) at base-level resolution (see Methods; Additional file 8: Figure S4<!--Supplementary Figure 2-->).

When trained using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb, respectively, the *preciseTAD* model predicted a total of 12,258 TAD and 15,707 chromatin loop boundaries in GM12878, as well as 9,603 TAD and 11,154 chromatin loop boundaries in K562 cell line (Additional file 9: Table S4<!--Supplementary Table 4-->). We reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Additional file 3: Table S2<!--Supplementary Table 3-->). This can be attributed to Arrowhead's inflation of called TADs at 5 kb, that, when visualized, often do not correspond to domains enriched in internal interactions (Figure 5A) and signal of known drivers of domain boundaries (Figure 5B). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Additional file 3: Table S2<!--Supplementary Table 3-->). This can be attributed to Peakachu's use of only CTCF sites to call boundaries, while *preciseTAD* leverages four known drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. These results suggest that genome annotation-driven predicted boundaries may be more biologically relevant than those detected from Hi-C data alone.

<!-- ![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of Arrowhead-called TAD boundaries (blue) vs. *preciseTAD*-predicted TAD boundaries (green) on GM12878 data (chr14:50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B) A zoomed-in portion of the genome shows the preciseTAD boundary region (PTBR, highlighted yellow), a cluster of bases with high probability of being a boundary, and the corresponding signal profiles of CTCF, RAD21, SMC3, and ZNF143. (C, D) Signal profile plots comparing the strength of CTCF, RAD21, SMC3, and ZNF143 binding around Arrowhead-called boundaries (blue, C), Peakachu loop boundaries (red, D) vs. *preciseTAD*-predicted boundaries (green).](Figures/figure5.png) -->

>>>>

---
title: "preciseTAD: A machine learning framework for precise 3D domain boundary prediction at base-level resolution"
csl: ../styles.ref/genomebiology.csl
output:
  pdf_document: 
    extra_dependencies: ["float"]
  word_document:
    reference_docx: ../styles.doc/Arial_11_double_space_normal_margins.docx
bibliography: ../References/references.bib
---

@@

<!--revision symbol: %%%-->

@@

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of the 3-dimensional (3D) structure of the human genome such as chromatin loops [@rao20143d], Topologically Associating Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At the kilobase scale, chromatin loops connect gene promoters with distal enhancers, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. At the megabase scale, TADs represent regions on the linear genome that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. Disruption of boundaries demarcating TADs and loops promotes cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Identifying the precise location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

@@

Another limitation among domain callers is that they disregard prior knowledge about functional genomic annotations associated with TAD and loop boundaries, referred hereafter as domain boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been identified as components of the loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Furthermore, distinct patterns of histone modifications have also been shown to be present at boundaries [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These genomic annotations are obtained using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). The resolution of ChIP-seq experiments is typically on the order of tens to hundreds of bases [@mendoza2009genome], well below the resolution of Hi-C data (tens of kilobases; 750 bp is the highest resolution of Hi-C data to date [@bonev2017multiscale]). Therefore, leveraging precisely mapped genomic annotations in a supervised machine learning framework enables more precise prediction of the locations of domain boundaries.

We present *preciseTAD*, an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on high-resolution chromatin state (BroadHMM), histone modification (HM), and transcription factor binding site (TFBS) data to predict both low-resolution TAD and chromatin-loop boundaries. We introduce a systematic pipeline for building the optimal boundary region prediction classifier. We found that spatial associations (linear distance) between boundaries and annotations perform best, transcription factor binding sites improve prediction performance, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and partitioning around medoids (PAM) to detect genome annotation-guided boundary regions and points at a base-level resolution. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that *preciseTAD* predictions are more enriched for known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD*-predicted boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in the domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on https://bioconductor.org/packages/preciseTAD.

@@

We developed a machine learning (ML) framework for determining the optimal set of data level characteristics to predict boundary regions of Topologically Associating Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We used Arrowhead-called TAD boundaries [@durand2016juicer] and published Peakachu-predicted loop boundaries [@salameh2019supervised] as ground truth. Data from GM12878 and K562 cell lines at 5-100 kb resolution (Arrowhead) and 10 kb resolution (Peakachu) were used (Additional file 1: Arrowhead Script, Additional file 2: Table S1). We next developed *preciseTAD*, an R package that leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base-level resolution from cell line-specific genome annotation data.

<!--![**Resolution-specific data construction and feature engineering for random forest modeling.** (A) The linear genome was binned into non-overlapping resolution-specific intervals using *shifted binning* (see Methods). The response vector **Y** was defined as 1/0 if a genomic bin overlapped/did not overlap with a TAD (or loop) boundary. (B) Four types of associations between bins (blue dashed lines) and genomic annotations (green shapes) were considered to build the predictor space, including Average Peak Signal (Signal), Overlap Counts (OC), Overlap Percent (OP), and $log_{2}$ distance (Distance).](../main_figures/figure1.png)-->

Boundary regions were defined as genomic bins containing a called boundary ($Y = 1$), while non-boundary regions were defined as bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Additional file 3: Table S2). The number of non-boundary regions highly outnumbered boundary regions. Such a disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address the class imbalance, we evaluated the effect of three resampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority class (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

|          Tool         |     Resolution/Bin Size    |     Total number of called TADs/chromatin loops    |     Total number of unique TAD/chromatin loop boundaries    |     Total number of genomic bins    |     Class imbalance    |
|:---------------------:|:--------------------------:|:--------------------------------------------------:|:-----------------------------------------------------------:|:-----------------------------------:|:----------------------:|
|        Arrowhead      |       5   kb               |                         8052                       |                             15468                           |                535363               |           0.03         |
|     Arrowhead         |       10   kb              |                         7676                       |                             14253                           |                267682               |           0.05         |
|     Arrowhead         |       25   kb              |                         4670                       |                             8363                            |                107073               |           0.08         |
|     Arrowhead         |       50   kb              |                         2349                       |                             4224                            |                 53537               |           0.08         |
|     Arrowhead         |      100   kb              |                         1031                       |                             1883                            |                 26768               |           0.07         |
|        Peakachu       |       10   kb              |                        16185                       |                          21421                              |                267682               |        0.14            |
|        Grubert        |       5    kb              |                        16232                       |                          18455                              |                535363               |        0.07            |
Table: Domain boundary data and class imbalance summaries across resolutions for Arrowhead, Peakachu<!--%%%, and Grubert--> on GM12878.

A total of 77 cell line-specific genomic annotations were used to build the predictor space. These included BroadHMM chromatin state data, histone modifications (HM), and transcription factor binding sites (TFBS) (Additional file 4: Table S3). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These included signal strength association (Signal), direct (OC), proportional (OP), and spatial ($log_2 + 1$ Distance) relationships (see Methods, Additional file 5: Figure S1).

@@

<!-- ![**A machine learning framework for building domain boundary region prediction models.** Step 1 employs a range of feature engineering techniques to define the predictor matrix $A_{N \times (p)}$, where $N$ is the number of genomic bins, $p$ is the number of genomic annotations, $i$ is a holdout chromosome. The response vector $Y_N$ is defined as a boundary region ($Y=1$) if it overlaps with a genomic bin (else $Y=0$). Step 2 reserves the predictor-response matrix for the holdout chromosome $i$ as the test data. Step 3 applies a resampling technique to the training data to address the class imbalance. Step 4 trains the random forest model and performs 3-fold cross-validation to tune the mtry parameter. Finally, step 5 validates the model on the separate test data composed of the binned data from the holdout chromosome $i$ and evaluates model performance using balanced accuracy (BA).](../main_figures/figure2.png) -->

<!--%%%
![**A machine learning framework for building domain boundary region prediction models.** Step 1 combines the response vector ($\textbf{Y}_N$) from *shifted binning* and a feature engineering procedure to form the data matrix $A_{N \times (p+1)}$ for chromosomes $\{1, 2, ..., i-1, i+1, ..., 22 \}$. Step 2 reserves the predictor-response matrix for the holdout chromosome $i$ as the test data using Grubert-defined boundaries as $\textbf{Y}_{test}$. Step 3 applies a resampling technique to the training data to address the class imbalance. Step 4 trains the random forest model and performs 3-fold cross-validation to tune the mtry parameter. Finally, step 5 validates the model on the separate test data composed of the binned data from the holdout chromosome $i$ and evaluates the model performance using balanced accuracy (BA)](../main_figures/figure2.png)
-->

@@

Additionally, we found that using a distance-type predictor space yielded substantially higher balanced accuracy than the peak signal strength, overlap count, and overlap percent predictor types. As with class balancing techniques, this improvement was less evident at lower resolutions, with results consistent for K562 (Additional file 6: Figure S2A). Furthermore, 5 kb resolution genomic bins led to the optimal prediction for TAD boundary regions on both cell lines. Random forest models built on Peakachu-defined loop boundary regions yielded optimal prediction performance when using a distance-type predictor space, with SMOTE resampling only marginally outperforming RUS, for both cell lines (Additional file 6: Figure S2B, S2C). Our results indicate that random under-sampling, distance-type predictors, and high-resolution Hi-C data provide the optimal set of data level characteristics for both TAD and chromatin loop boundary prediction.

<!-- ![**Determining optimal data level characteristics for building TAD boundary region prediction models on GM12878.** Averaged balanced accuracies are compared across resolution, within each predictor-type: Signal, OC, OP, and Distance, and across resampling techniques: no resampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Error bars indicate 1 standard deviation from the mean performance across each holdout chromosome used for testing.](../main_figures/figure3.png) -->

@@

We hypothesized that the type of genomic annotations may also affect predictive performance. Using the established optimal settings (RUS, Distance, 5 kb/10 kb (Arrowhead/Peakachu ground truth) genomic bins), we built separate random forest models using histone modifications (HM), BroadHMM chromatin states (BroadHMM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, with results consistent for loop boundaries, on both cell lines (Figure 4A; Additional file 7: Figure S3A). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both GM12878 and K562.

<!-- ![**SMC3, RAD21, CTCF, and ZNF143 transcription factors accurately predict TAD and loop boundaries in GM12878.** (A) Barplots comparing performances of TAD (Arrowhead) and loop (Peakachu) boundary prediction models using histone modifications (HM), chromatin states (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent 1 standard deviation from the mean cross-validated accuracy across each holdout chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 most predictive chromosome-specific TFBSs. The columns represent the holdout chromosome excluded from the training data. Rows are sorted in decreasing order according to the columnwise average importance. (D) Receiver operating characteristic (ROC) curves and the corresponding average area under the curves (AUCs) when training and testing on GM12878 data (blue, Arrowhead ground truth; red, Peakachu ground truth) versus training on K562 and testing on GM12878 data (black, dashed). The curves represent the average sensitivities and specificities across each holdout chromosome. The shaded areas around each curve represent 1 standard deviation from the average.](../main_figures/figure4.png) -->

@@

We sought to further optimize our boundary region prediction models by implementing recursive feature elimination to avoid overfitting and selecting only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Additional file 7: Figure S3B). However, given that we trained our models on chromosome-specific data, the most significant annotations varied for each holdout chromosome. To determine transcription factors most important for boundary prediction across all chromosomes, we clustered the predictive importance (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found four transcription factors, CTCF, RAD21, SMC3, and ZNF143, being consistently predictive of TAD and loop boundaries (Figure 4C; Additional file 7: Figure S3C). We optimized our model by only considering these top four TFBS when building the random forest model, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the known molecular drivers of the loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line using the model pre-trained on data from another cell line (See Methods). We found that training and testing using Arrowhead ground truth TAD boundaries and genomic annotation data from the GM12878 cell line resulted in an average AUC=0.792 (Figure 4D). Interestingly, when training on the K562 cell line and testing on GM12878, the average AUC increased slightly to 0.795. Likewise, the average performance of models trained using Peakachu boundaries and genomic annotation data from the GM12878 cell line was comparable to models trained on K562-specific Peakachu boundaries and genomic annotations (Avg. AUC=0.881 and 0.874, respectively). These results were consistent when comparing training/testing strategies on K562 with training on GM12878 and testing on K562 data (Additional file 7: Figure S3D). The average ROC curves were found to be within 1 standard deviation of each other, suggesting that a model trained on data from one cell line performs well when using the data from another cell line, allowing for the opportunity to predict boundaries for cell lines that do not currently have Hi-C data available.

@@

Using our optimally built random forest model trained on Arrowhead/Peakachu boundaries, we attempted to predict boundaries at base-level resolution. Intuitively, instead of bin-level annotations, the predictor-response space was built on a base-level, that is, each base was associated with a genomic annotation. The model trained on a bin-level space was then applied on a base-level space to predict each base's probability of being a boundary. Our method, referred to as *preciseTAD*, uses density-based spatial clustering (DBSCAN) and partitioning around medoids (PAM) to predict the location of boundary regions and summit points (highest boundary probability) at base-level resolution (see Methods; Additional file 8: Figure S4).

When trained using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb, respectively, the *preciseTAD* model predicted a total of 12,258 TAD and 15,707 chromatin loop boundaries in GM12878, as well as 9,603 TAD and 11,154 chromatin loop boundaries in K562 cell line (Additional file 9: Table S4). We reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Additional file 3: Table S2). This can be attributed to Arrowhead's inflation of called TADs at 5 kb, that, when visualized, often do not correspond to domains enriched in internal interactions (Figure 5A) and signal of known drivers of domain boundaries (Figure 5B). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Additional file 3: Table S2). This can be attributed to Peakachu's use of only CTCF sites to call boundaries, while *preciseTAD* leverages four known drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. These results suggest that genome annotation-driven predicted boundaries may be more biologically relevant than those detected from Hi-C data alone.

<!-- ![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of Arrowhead-called TAD boundaries (blue) vs. *preciseTAD*-predicted TAD boundaries (green) on GM12878 data (chr14:50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B) A zoomed-in portion of the genome shows the preciseTAD boundary region (PTBR, highlighted yellow), a cluster of bases with high probability of being a boundary, and the corresponding signal profiles of CTCF, RAD21, SMC3, and ZNF143. (C, D) Signal profile plots comparing the strength of CTCF, RAD21, SMC3, and ZNF143 binding around Arrowhead-called boundaries (blue, C), Peakachu loop boundaries (red, D) vs. *preciseTAD*-predicted boundaries (green).](../main_figures/figure5.png) -->

<!--%%%
![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of Arrowhead-called TAD boundaries (blue) vs. *preciseTAD*-predicted TAD boundaries (green) on GM12878 data (chr14:50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B) A zoomed-in portion of the genome shows the preciseTAD boundary region (PTBR, highlighted yellow), a cluster of bases with high probability of being a boundary, and the corresponding signal profiles of CTCF, RAD21, SMC3, and ZNF143.](../main_figures/figure5.png)
-->

Next, we evaluated the biological significance of *preciseTAD* boundary points. The signal of four known molecular drivers of 3D chromatin (CTCF, RAD21, SMC3, and ZNF143) colocalized more frequently around *preciseTAD*-predicted boundary points, as compared to Arrowhead-called TAD and Peakachu loop boundaries, respectively (Figure 5C, 5D; Additional file 10: Figure S5A, S5B). Surprised by the poor signal distribution around Arrowhead boundaries, we compared signals centered on boundaries called by Arrowhead, Peakachu, and our SpectralTAD [@Cresswell:2020ab] TAD caller. We confirmed the poor signal distribution around Arrowhead boundaries, in contrast to the relatively well-performing Peakachu- and SpectralTAD-called boundaries (Additional file 10: Figure S5C, S5D). Signal enrichment heatmaps confirmed that *preciseTAD*-predicted boundaries were more enriched for the same genomic annotations than either Arrowhead or Peakachu boundaries alone (Additional file 11-12: Figure S6-S7). *preciseTAD* boundaries were statistically significantly closer to the top-ranked TFBS (Wilcoxon p-value < 0.001 versus Arrowhead and Peakachu boundaries, respectively, Figure 6A, Additional file 13: Figure S8). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation.
